---
title: "EDA"
author: "Helen Lan & Rachel Yang"
date: "4/7/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(broom)
#library(tidylog)
library(ggfortify)
library(survival)
library(reshape2)
library(ggplot2)
```

```{r, include=FALSE}
AIDS <- read_csv("~/Desktop/Math150-Project/AIDSdata.csv")
```
```{r}
summary(AIDS$age)
summary(AIDS$time_d)
AIDS$SurvObj <- with( AIDS, Surv(time_d, censor == 0))
km.AIDS <- survfit(SurvObj ~ 1, data = AIDS, conf.type = "log-log")
autoplot(km.AIDS)
```
Something New
1. Investigation of PH assumption (Rachel)
1.1 Introduction 
1.1.1 What is PH Assumption, and why it is important?
    Proportional Harzard(PH) Assumption is the assumption that any two groups 
    in the study must have hazard functions that are proportional over time 
    (i.e. have parallel curve). A non-proportional hazard means that there is 
    an interation between our independent variable and time, therefore we need 
    to do more work and be more careful when designing our model.
1.1.2 Resources
    My knowledge of the concept of PH Assumption mainly comes from 
    class lectures. I will use r documentation(linked below) to learn about the 
    cox.zph function.
    https://www.rdocumentation.org/packages/survival/versions/2.43-3/topics/cox.zph
1.1.3 Challenges
    The first challenge is figuring out the r code. The second challenge is 
    interpretating of the result and deciding what to do with our model base on 
    the result.
1.2 Testing the PH Assumption
1.3 Interpretation
1.4 Actions
